Glacial flows, simulated faster

### 2023-06-05 talk-iu/slides.tex
edits to later slides ... I think only a couple of slides on multilevel VI method are needed

\medskip
\item<2> progress has been made on well-posedness of the lubrication approximation of the basic model, the so-called \alert{shallow ice approximation}:
    \begin{itemize}
    \item[$\circ$] 1D well-posedness on flat bed (Calvo et al 2003)
    \item[$\circ$] 2D steady-state existence on general beds (Jouvet \& Bueler 2012)
    \item[$\circ$] 2D well-posedness on flat bed (Piersanti \& Temam 2022)
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{explicit ice sheet model performance}

@@

\vspace{10mm}
\begin{itemize}
\item explicit time-stepping implies \alert{many stress-balance solves}
    \begin{itemize}
    \item[$\circ$] while stress-balance scaling exponent $\alpha$ is important, even optimality ($\alpha=0$) cannot rescue performance
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item switch to \alert{implicit time-stepping} for unconditional stability?
    \begin{itemize}
    \item[$\circ$] each step is a \alert{free-boundary} NCP-coupled-to-Stokes problem
    \item[$\circ$] parameterize cost of these solves as $O(m^{1+\beta})$
    \end{itemize}
\item need $q$ model updates per year to integrate climate influences
\end{itemize}
\end{frame}

@@

\bigskip
\begin{itemize}
\item goal for optimists: implicit time-stepping \emph{and} build a $\beta \approx 0$ NCP-coupled-to-Stokes solver for the time step problems
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item no convincing NCP-coupled-to-Stokes (free-boundary) solvers exist yet
    \begin{itemize}
    \item[$\circ$] Wirbel \& Jarosch~(2020) is an important attempt \dots
    \end{itemize}
\item the Bueler (2016) implicit (free-boundary) SIA solver scales badly: $\beta=0.8$
\end{itemize}
\end{frame}

\section{a multilevel approach to better performance}

@@

\begin{itemize}
\item ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
   \begin{itemize}
   \item[$\circ$] there are \alert{no easy or magic techniques} for performance
   \end{itemize}
\item<2-> current-technology ice sheet models mostly use \alert{explicit} time stepping, \alert{non-optimal} stress-balance solvers, and \alert{shallow} assumptions
   \begin{itemize}
   \item[$\circ$] progress is being made in all of these areas, e.g.~ scalable Stokes solvers (Isaac et al.~2015)
   \end{itemize}
\item<3> scalable solvers for implicit-step NCP-coupled-to-Stokes models, which would seem to be the recommended numerical design, require \alert{multilevel solvers for non-local variational inequalities}
\end{itemize}
\end{frame}

\begin{frame}{references}

{\scriptsize
\begin{itemize}
\item E.~Bueler (2016). \emph{Stable finite volume element schemes for the shallow-ice approximation}, J.~Glaciol.~62 (232), 230--242, \href{https://doi.org/10.1017/jog.2015.3}{10.1017/jog.2015.3}
\item E.~Bueler (2021). \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math.~81 (5), 2007--2032, \href{https://doi.org/10.1137/20M135217X}{10.1137/20M135217X}
\item E.~Bueler (2022). \emph{Performance analysis of high-resolution ice-sheet simulations}, J.~Glaciol., \href{https://doi.org/10.1017/jog.2022.113}{10.1017/jog.2022.113}
\item E.~Bueler \& P.~Farrell (in preparation). \emph{A full approximation storage multilevel method for nonlinear variational inequalities}
% Calvo et al 2002
\item T.~Isaac, G.~Stadler, \& O.~Ghattas (2015). \emph{Solution of nonlinear Stokes equations discretized by high-order finite elements on nonconforming and anisotropic meshes, with application to ice sheet dynamics}, SIAM J.~Sci.~Comput.~37 (6), B804--B833, \href{https://doi.org/10.1137/140974407}{10.1137/140974407}
\item G.~Jouvet \& E.~Bueler (2012). \emph{Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation}, SIAM J.~Appl.~Math.~72 (4), 1292--1314, \href{https://doi.org/10.1137/110856654}{10.1137/110856654}
\item P.~Piersanti \& R.~Temam (2022). \emph{On the dynamics of grounded shallow ice
sheets: modeling and analysis}, Adv.~Nonlinear Analysis 12, \href{https://doi.org/10.1515/anona-2022-0280}{10.1515/anona-2022-0280}
% Shapero et al ?
% Wirbel & Jarosch
\end{itemize}
}
\end{frame}

>>>>

\medskip
\item<2> progress has been made on well-posedness of the lubrication approximation of the basic model, the so-called \alert{shallow ice approximation}:
    \begin{itemize}
    \item[$\circ$] 1D well-posedness on flat bed (Calvo et al 2002)
    \item[$\circ$] 2D steady-state existence on general beds (Jouvet \& Bueler 2012)
    \item[$\circ$] 2D well-posedness on flat bed (Piersanti \& Temam 2022)
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{ice sheet models: explicit time-stepping performance}

@@

\vspace{10mm}
\begin{itemize}
\item we \emph{want} optimality: \quad $O(m^1)$ flops per model year
\item explicit time-stepping implies \alert{too many stress-balance solves}
    \begin{itemize}
    \item[$\circ$] while the Stokes (stress-balance) scaling exponent $\alpha$ is important, even Stokes solver optimality ($\alpha=0$) cannot yield optimality
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item let us try \alert{implicit time-stepping}, for its unconditional stability
\item each step is now a \alert{free-boundary NCP-coupled-to-Stokes problem}
\item let us parameterize cost of these solves as $O(m^{1+\beta})$
\item we still need $q$ model updates per year to integrate climate influences, and track evolution for the simulation purpose
\end{itemize}
\end{frame}

@@

\bigskip
\begin{itemize}
\item new goal: use implicit time-stepping \emph{and}~build a $\beta \approx 0$ NCP-coupled-to-Stokes solver for problem at each time step
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item no convincing NCP-coupled-to-Stokes (free-boundary) solvers exist yet
    \begin{itemize}
    \item[$\circ$] however, Wirbel \& Jarosch~(2020) is an important beginning
    \end{itemize}
\item the Bueler (2016) implicit and NCP SIA solver scales badly: $\beta=0.8$
\end{itemize}
\end{frame}

\section{a multilevel approach}

@@

\begin{itemize}
\item glacier simulations are both \alert{important to humanity} and a rich \alert{source of interesting mathematics}
   \begin{itemize}
   \item[$\circ$] predict sea level rise!
   \end{itemize}
\item<2-> ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
   \begin{itemize}
   \item[$\circ$] there are \alert{no easy or magic techniques} for performance
   \end{itemize}
\item<3-> current-technology ice sheet models mostly use \alert{explicit} time stepping, \alert{non-optimal} stress-balance solvers, and \alert{shallow} assumptions
   \begin{itemize}
   \item[$\circ$] progress is being made in all of these areas, e.g.~scalable Stokes solvers (Isaac et al.~2015)
   \end{itemize}
\item<4> scalable solvers for implicit-step, NCP-coupled-to-Stokes models require \alert{multilevel solvers for non-local variational inequalities}
   \begin{itemize}
   \item[$\circ$] is this the preferred numerical design for the basic glacier model?
   \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{references (partial list)}

{\scriptsize
\begin{itemize}
\item E.~Bueler (2016). \emph{Stable finite volume element schemes for the shallow-ice approximation}, J.~Glaciol.~62 (232), 230--242, \href{https://doi.org/10.1017/jog.2015.3}{10.1017/jog.2015.3}
%\item E.~Bueler (2021). \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math.~81 (5), 2007--2032, \href{https://doi.org/10.1137/20M135217X}{10.1137/20M135217X}
\item E.~Bueler (2022). \emph{Performance analysis of high-resolution ice-sheet simulations}, J.~Glaciol., \href{https://doi.org/10.1017/jog.2022.113}{10.1017/jog.2022.113}
\item E.~Bueler \& P.~Farrell (in preparation). \emph{A full approximation storage multilevel method for nonlinear variational inequalities}
\item N.~Calvo, J.~D\'iaz, J.~Durany, E.~Schiavi, \& C.~V\'azquez (2002). \emph{On a doubly-nonlinear parabolic obstacle problem modelling ice sheet dynamics}, SIAM J.~Appl.~Math., 63(2), 683--707 \href{https://doi.org/10.1137/S0036139901385345}{10.1137/S0036139901385345}
\item T.~Isaac, G.~Stadler, \& O.~Ghattas (2015). \emph{Solution of nonlinear Stokes equations discretized by high-order finite elements on nonconforming and anisotropic meshes, with application to ice sheet dynamics}, SIAM J.~Sci.~Comput.~37 (6), B804--B833, \href{https://doi.org/10.1137/140974407}{10.1137/140974407}
\item G.~Jouvet \& E.~Bueler (2012). \emph{Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation}, SIAM J.~Appl.~Math.~72 (4), 1292--1314, \href{https://doi.org/10.1137/110856654}{10.1137/110856654}
\item P.~Piersanti \& R.~Temam (2022). \emph{On the dynamics of grounded shallow ice
sheets: modeling and analysis}, Adv.~Nonlinear Analysis 12, \href{https://doi.org/10.1515/anona-2022-0280}{10.1515/anona-2022-0280}
\item A.~Wirbel \& A.~Jarosch (2020). \emph{Inequality-constrained free-surface evolution in a full Stokes ice flow model (evolve\_glacier v1.1)}, Geosci.~Model Dev.~13 (12), 6425--6445, \href{https://doi.org/10.5194/gmd-13-6425-2020}{10.5194/gmd-13-6425-2020}
\end{itemize}
}
\end{frame}
